Two Dimensional CSDA at the scale of neuronal tissue.

### 2015-07-01 Reportes/CSDAattheSomaticLayersScale.tex
Pum pum CSDA

\title{CSDA in 2D slices of the hippocampus \emph{done right}}
\author{W. P. K. Zapfe, Whoever Helps}

@@

\begin{abstract}
As high density electrode arrays become cheaper and more common, 
electrophysiological measurements of activity in neural tissue approach
the resolution of the studied structures. The electrodes measure
directly the Local Field Potential, but it is aknowledged that the
Current Source Density has more resolving power, as it points
more accurately to the actual sources and sinks themselves,
and thus, to the more active places in the tissue. Nevertheless,
the role of the conductivity tensor is rarely taken into account
in the derivation of the CSD, which could potentialy lead to
spurious distributions of the sources and sinks and deformation
of their structure, obfuscating the precise place of the
poles underneath.
Here we show a possible qualitative treatment of the conductivity
across the stratum piramidale of the hippocampus, an examplification
with two toy models and a possible numerical implementation with
high density experimental data.
\end{abstract}

@@

As Bédard and Destexhe pointed out in \cite{Bedard11}, the CSDA, as it
is performed usually, rests on
assumptions which could leave important features out. One of these 
assumptions is that the extracellular medium has very homogeneous 
electrical propierties, but this has been shown to not be
the case for dense neuronal layers such as the piramidal
strata in the hippocampus \cite{Holsheimer87, Lopez01, TrevinoPersonal}.
The evidence mostly points out that resistivity grows 
considerably around thick somatic layers, or conversely, that
the conductivity drops considerably there. In ohmic (linear) media,
we expect the following relation to hold:
\begin{equation}
\Jd=\cond \EF,
\end{equation}
that is, the current density $\Jd$ is the result of the
local electric field $\EF$ and the local conductivity, $\cond$.
We usually take that last quantity as an homogeneous, isotropic,
factor, at least over the scales that we are interested in.
The conductivity in inhomogeneous media can be rarely
simplified to a constant, as it can depend both from the direction
of the fields and currents, and vary from point to point. 
Then the above expression is just shorthand notation for the
explicit  \emph{tensorial} relationship:
\begin{equation}
  \begin{pmatrix}
    J_x \\
    J_y \\
    J_z
  \end{pmatrix}
  =
   \begin{pmatrix}
     \sigma_{xx}E_x+ \sigma_{xy}E_y+\sigma_{xz}E_z \\
     \sigma_{xy}E_x+ \sigma_{yy}E_y+\sigma_{yz}E_z \\
     \sigma_{xz}E_x+ \sigma_{yz}E_y+\sigma_{zz}E_z \\
  \end{pmatrix},
\end{equation}
where we allready used the fact that $\cond$ is simmetric when
expressed on cartessian coordinates.
We assume that the electric field can be obtained as the
gradient of the LFP, $\phi$:
\begin{equation}
  -\grad \phi=\EF
\end{equation}
This LFP is the quantity that the microelectrode arrays (MEAs) permits us
to measure on the extracelular medium.
To obtain the sources and sink densities, we use Gauss Divergence Theorem,
stated in differencial form:
\begin{equation}
\nabla \cdot \Jd =I_m
\end{equation}
We perform the divergence operation on the current density, not
on the electric field. Only when the medium is isotropic and homogeneous
we can be sure that both operations coincide within a positive factor.

Here we shall investigate
posible influence of changing conductivity on current density
($\Jd$) and current source density ($I_m$).

@@

In experiments of this sort performed on hippocampus slices, the
MEAs permit us to measure a sample of LFP over discrete points on a
two dimensional lattice, so we do not have possibility of
calculating currents or electric fields on the third dimension of space.
The slices of the hippocampus are very thin, and we do not consider
the posibility of current going outside the neural tissue sample (in one
direction there is air and on the other the MEA), so we
can reduce our formalism to two dimensional space. We then take
$J_z=0$ and use $(x,y)$ as a cartesian coordinate system aligned
with the MEA. Then the \emph{flat} current density  is:
\begin{equation}
  \begin{pmatrix}
    J_x\\
    J_y
  \end{pmatrix}  =
  \begin{pmatrix}
    \sigma_{xx}E_x+\sigma_{xy}E_y \\
    \sigma_{xy}E_x+\sigma_{yy}E_y 
  \end{pmatrix}=
  \begin{pmatrix}
    -\sigma_{xx}\partial_x \phi -\sigma_{xy}\partial_y \phi \\
    -\sigma_{xy}\partial_x \phi -\sigma_{yy}\partial_y \phi 
  \end{pmatrix}.
\end{equation}
The divergence, in cartesian coordinates, yields:
\begin{equation}\label{csd2dexp}
  I_m=\partial_x J_x+\partial_y J_y=
  -\partial_x( \sigma_{xx}\partial_x \phi +\sigma_{xy}\partial_y \phi)
  -\partial_y( \sigma_{xy}\partial_x \phi +\sigma_{yy}\partial_y \phi). 
\end{equation}
Each one of the terms of the right hand side of eq. \ref{csd2dexp}
has the form
\begin{multline}
  \partial_j (\sigma_{jj}\partial_j\phi) +\partial_j (\sigma_{jk}\partial_k \phi)
  =  \\
  (\partial_j \sigma_{jj})(\partial_j\phi) +\sigma_{jj}\partial_j^2\phi +
  (\partial_j \sigma_{jk})(\partial_k \phi)+\sigma _{jk}\partial_j\partial_k \phi.
\end{multline}
The complete expression is cumbersome to write, but it is important to
have \emph{full} expansion in order to make the numerical considerations
in what follows.
\begin{equation}\label{Imcart}
  \begin{split}
  -I_m= & (\partial_x \sigma_{xx}) (\partial_x \phi) +
  (\partial_y \sigma_{yy}) (\partial_y \phi)\\
& +(\partial_x \sigma_{xy}) (\partial_y \phi) +(\partial_y \sigma_{xy}) (\partial_x \phi)\\
& + \sigma_{xx} \partial_x^2 \phi + \sigma_{yy} \partial_y^2 \phi\\
& + 2\sigma_{xy} \partial_x\partial_y \phi. 
\end{split}
 \end{equation}

@@

This expressions may be regardes as inescesarily long, as the 
$\cond$ tensor, being a real symmetric tensor, can be represented
in local coordinates which diagonalize it. Those coordinates have
physical meaning, as they are the resolving directions in which
conductivity behaves differently and denote some underlying structure
in the studied material. The usual assumption is that 
the conductivity tensor is isotropic but not homogeneous, and it 
changes along the apical direction, and not along the 
differentiated strata (e.g. stratum pyramidale, stratum granolosum).
Let us suppose that this is approximately correct, and that very close
to the stratum pyramidale we represent $\cond$ into the coordinates
defined by its eigenvectors (see fig. \ref{esquemas01}).
 We shall call them $(t,a)$ coordinates
and asume that the first is approximately tangent to the
strata and the other to the apical direction. The $\cond$ tensor would 
have the expression in such coordinates
\begin{equation}
\tilde{\cond}=
\begin{pmatrix}
\sigma_t & 0 \\
0 & \sigma_a
\end{pmatrix}
\end{equation} 
Then, the equation \ref{Imcart} becomes
\begin{equation}\label{Imapic}
\begin{split}
-I_m= & (\partial_t \sigma_{t}) (\partial_t \phi) +
(\partial_a \sigma_{a}) (\partial_a \phi)  \\
& + \sigma_{a} \partial_a^2 \phi + \sigma_{t} \partial_t^2 \phi, 
\end{split}
\end{equation}
as $(t,a)$ are also local ortogonal coordinates. 
One may wonder why we are explicitly stating the non-diagonal 
version of the equations. The data as it is delivered by
the measuring apparatus is on a strictly cartessian grid, and
we have to perform the numerical derivatives over this data, so
we have to know how these assumptions appear in both sets
of coordinates. Besides, the \emph{tangent-apical} coordinates
are just a teoretical tool, a numerical implementation of such a coordinate
set would be  complicated and prone to numerical errors.

>>>>

\title{Two Dimensional CSDA at the scale of neuronal tissue.}
\author{W. P. K. Zapfe, Whoever Helps}

@@

\begin{abstract}
As high density electrode arrays become cheaper and more common, 
electrophysiological measurements of activity in neural tissue approach
the scale of the underlying structures. The electrodes measure
directly the Local Field Potential, but it is aknowledged that the
Current Source Density has more resolving power and distinguishes
clearly between the loci of outgoing current and ingoing currents.
 The role of the conductivity tensor is rarely taken into account
in the derivation of the CSD, and this could potentialy lead to
spurious distributions of the sources and sinks- 
In dense neural tissue, such as the
stratum piramidale layer of the hippocampus, clear
deviations from average conductivity have been measured. 
Here we show a possible qualitative treatment of the conductivity
across the layers of the hippocampus, and compare the results
when this inhomogenity is taken into account and when it not.
\end{abstract}

@@

As Bédard and Destexhe pointed out in \cite{Bedard11}, the CSDA, as it
is performed usually, rests on
assumptions which could leave important features out. One of these 
assumptions is that the extracellular medium has very homogeneous 
electrical propierties. This has been shown to not be
the case for dense neuronal layers such as the piramidal
strata in the hippocampus \cite{Holsheimer87, Lopez01, TrevinoPersonal}.
The evidence mostly points out that resistivity grows 
considerably around thick somatic layers, or conversely, that
the conductivity drops considerably there. In ohmic (linear) media,
we expect the following relation to hold:
\begin{equation}
\Jd=\cond \EF,
\end{equation}
that is, the current density $\Jd$ is the result of the
local electric field $\EF$ and the local conductivity, $\cond$.
We usually take that last quantity as an homogeneous, isotropic,
factor, at least over the scales that we are interested in.
The conductivity in inhomogeneous media can be rarely
simplified to a constant, as it can depend both from the direction
of the fields and currents, and vary from point to point. 
Then the above expression is just shorthand notation for the
explicit  \emph{tensorial} relationship:
\begin{equation}
  \begin{pmatrix}
    J_x \\
    J_y \\
    J_z
  \end{pmatrix}
  =
   \begin{pmatrix}
     \sigma_{xx}E_x+ \sigma_{xy}E_y+\sigma_{xz}E_z \\
     \sigma_{xy}E_x+ \sigma_{yy}E_y+\sigma_{yz}E_z \\
     \sigma_{xz}E_x+ \sigma_{yz}E_y+\sigma_{zz}E_z \\
  \end{pmatrix},
\end{equation}
where we allready used the fact that $\cond$ is simmetric when
expressed on Cartessian coordinates.
We assume that the electric field can be obtained as the
gradient of the LFP, $\phi$:
\begin{equation}
  -\nabla \phi=\EF
\end{equation}
This LFP is the quantity that the microelectrode arrays (MEAs) permits us
to measure on the extracelular medium.
To obtain the sources and sink densities, we use Gauss Divergence Theorem,
\begin{equation}
\nabla \cdot \Jd =I_m
\end{equation}
The usual definition of $-\partial_t \rho =I_m$ has been made here. 
We perform the divergence operation on the current density, \emph{not}
on the electric field. When the medium is isotropic and homogeneous
both derivations coincide within a positive factor. 
In the next section we investigate the role of the inhomogenities 
of the conductivity on current density
($\Jd$) and current source density ($I_m$) at scales around
the $20\mu m$.

@@

In experiments of this sort performed on hippocampus slices, the
MEAs permit us to measure a sample of LFP over discrete points on a
two dimensional lattice, so we do not have possibility of
calculating currents or electric fields on the third dimension of space.
The slices of the hippocampus are very thin, and we do not consider
the posibility of current going outside the neural tissue sample (in one
direction there is air and on the other the MEA), so we
can reduce our formalism to two dimensional space. We then take
$J_z=0$ and use $(x,y)$ as a cartesian coordinate system aligned
with the MEA. Then the \emph{flat} current density  is:
\begin{equation}
  \begin{pmatrix}
    J_x\\
    J_y
  \end{pmatrix}  =
  \begin{pmatrix}
    \sigma_{xx}E_x+\sigma_{xy}E_y \\
    \sigma_{xy}E_x+\sigma_{yy}E_y 
  \end{pmatrix}=
  \begin{pmatrix}
    -\sigma_{xx}\partial_x \phi -\sigma_{xy}\partial_y \phi \\
    -\sigma_{xy}\partial_x \phi -\sigma_{yy}\partial_y \phi 
  \end{pmatrix}.
\end{equation}
The divergence, in cartesian coordinates, yields:
\begin{equation}\label{Imcart}
  \begin{split}
  -I_m= & (\partial_x \sigma_{xx}) (\partial_x \phi) +
  (\partial_y \sigma_{yy}) (\partial_y \phi)\\
& +(\partial_x \sigma_{xy}) (\partial_y \phi) +(\partial_y \sigma_{xy}) (\partial_x \phi)\\
& + 2\sigma_{xy} \partial_x\partial_y \phi \\
& + \sigma_{xx} \partial_x^2 \phi + \sigma_{yy} \partial_y^2 \phi.
\end{split}
 \end{equation}
The \emph{tacit} assumption that can be noted in most of the CSDA previously
done is that the derivatives of $\cond$ are negligible and only
the terms on the last line are  taken into account. 
It can be argued that for scales that
encompas many neurons and interneuronal space, the averaged conductivity is
constant and has then its derivatives vanish. For MEAS that take measuraments
more than $100 \mu m$ appart, this may be a reasonable assumption (see
for example, eq. 12 in \cite{Bedard11} and the arguments leading to it).
Examples of such derivations are those obtained by the MED64 probe in the
work of Wang et all \cite{Wang10}, etc etc. Also sofisticated
numerical methods, such as iCSD \cite{Pettersen2006, Leski11}
and kCSD \cite{Potworowsky2011} use this approximation, albeit
explicitly.

If the interelectrode spacing
is less than $100 \mu m$ then the neuronal tissue would begin
to play a role in the inhomogenities of $\cond$, although the isotropic
assumption may still be valid at scales larger than $20 \mu m$. At such 
scales the laminar structure of the Cornu Amonis or the Gyrus Dentate is
the main feature, and we should expect variations on the electrical 
propierties of the media along the apical direction. Measurements made
by Holsheimer showed that measurements in the CA1 region of a mouse
hipocampus show variations of  $40\% $ in the resistivity around the mean
value at distances of about $60 \mu m$.
\cite{Holsheimer87}. This measurements have been confirmed
recently by a collaborator \cite{TrevinoPersonal}. On the other hand
we expect very little variation of the conductivity tensor along
each structural layer. The symmetry of the conductivity tensor represents
the structural underlying propierties of the media: as it can be 
represented on a diagonal form on a suitable set of \emph{orthogonal}
coordinates, then such coordinates must be associated with the 
structures that give $\cond$ its values. Then the representation
of $\cond$ in diagonal form should represent the conductivity
along the axonal-apical direction, with high variation, and another
axis transverse to it, in which $\cond$ showing very little variation.
We can imagine that we can set a locally orthonormal set of coordinates
alignate with such structures (see fig. \ref{esquemas01}).
 Then the expression in eq. \ref{Imcart}
can be reduced substantialy.  The $\cond$ tensor would appear as
\begin{equation}
\tilde{\cond}=
\begin{pmatrix}
\sigma_t & 0 \\
0 & \sigma_a
\end{pmatrix}
\end{equation} 
where $(t,a)$ are the transveral and axonal-apical local coordinates.
The CSD would be obtained as
\begin{equation}\label{Imapic}
\begin{split}
-I_m= & (\partial_t \sigma_{t}) (\partial_t \phi) +
(\partial_a \sigma_{a}) (\partial_a \phi)  \\
& + \sigma_{a} \partial_a^2 \phi + \sigma_{t} \partial_t^2 \phi. 
\end{split}
\end{equation}

@@
